Formal Grammars as Types in Non-Commutative Linear-Non-Linear Type Theory

### 2024-07-11 POPL/paper.tex
some comments

\steven{The above paragraph maybe isn't the right thing to say, but I want
  something that ties the first paragraph to the contributions that follow}
\pedro{As a rule of thumb, I think that it's good to avoid subjective and
  pointed adjectives such as ``intuitive'', espcially if we don't have evidence
  that corroborates our claim. This is the perfect excuse for
an annoying reviewer to complain about our work.}

@@

for every string $w \in \String$. Intuitively, we read a morphism of grammars as
a \emph{parse transformer} --- i.e. $f$ translates a $G$-parse of $w$ into an $H$-parse
of $w$. The induced notion of isomorphism in $\Grammar$ encodes that two grammars are equivalent if there is
a bijective translation of parses. In other words, two grammars are isomorhpic
if they capture the same structural descriptions of strings, which is
precisely Chomsky's notion of \emph{strong equivalence} of
grammars \cite{chom1963}. In works like
\cite{yoshinaga2002formal}, this notion of strong equivalence is often treated
like the appropriate choice of ``sameness'' for two grammars. When abstracting over the latent
structure of formal grammars and describing it in the the language of
categories, it is encouraging to see the same definition of equivalence arise
naturally.

Upon inspection of the above definition, $\Grammar$ may equivalently be described as the functor
category $\Set^{\String}$ --- where the set $\String$ is viewed as a
discrete\footnote{Objects in $\String$ comprise the set of all strings, and the
  only morphisms are the identity morphisms.}
category. In fact, in this paper we will take $\Grammar$ to be defined as the
functor category $\Set^{\String}$. Such functor categories into $\Set$ --- often called
presheaf categories --- carry a remarkable amount of structure.

@@

\steven{I initially wanted to say ``universal semantic domain to reason about
  parsing'', but these words have specific domain theory usage that aren't
  appropriate here, right?}
\pedro{If you want to keep ``semantic domain'', we could go for ``all-encompassing semantic domain''}
\steven{Check details for 2-cat of grammar formalisms}

@@

Omission of the structural rules of a deductive system, such as in
linear logic \cite{GIRARD19871}, offers precise control over how a value is used
in a derivation. Namely, linear logic omits the weakening and contraction rules
to ensure that every value in context is used exactly once. This control enables
\emph{resource-sensitive} reasoning, where we may treat a resource as
\emph{consumed} after usage. This viewpoint is amenable to parsing applications,
as we may treat characters of a string as finite resources that are consumed at
parse-time. That is, once a particular substring has been parsed, a parser need not go
over it again. One may then envision a linear type system where the types comprise
formal grammars generated over some alphabet $\Sigma$, and the type constructors
correspond precisely to inductive constructions on grammars --- such as
conjunction, disjunction, concatenation, etc.

Unfortunately, programming in a linear term language is often
cumbersome and unintuitive for users. Code in such a language can
become unnecessarily verbose when using Girard encodings to translate
high-level reasoning into low-level linear terms.\pedro{What encodings are these?
We should add a citation here}
To alleviate this
concern, in 1995 Benton et al.\ proposed an alternative categorical
semantics of linear logic based on adjoint interactions between linear
and non-linear logics \cite{bentonMixedLinearNonlinear1995} ---
appropriately referred to as a \emph{linear-non-linear} (LNL)
system. This work is simply typed, so the boundary between linear and
non-linear subtheories is entirely characterized via a monoidal
adjunction $F \dashv G$ between linear terms and non-linear terms.

@@

The structural judgments are shown in \cref{fig:structjdg}, the typing
well-formedness in \cref{fig:typewf}, and the intuitionistic typing
rules in \cref{fig:inttyping}. These are mostly just as they appear in
\cite{krishnaswami_integrating_2015}. It has two main differences. The
first one it the presence of two-distinct arrow types, one for adding
variables to the beginning of contexts and another one for adding
variables to the end of contexts. This is an adequate change in the
context of grammars, which are inherently non-symmetric. The second
change is the introduction of inductive types, which also have a
natural grammar counterpart, which usually allows kinds of recursive
behavior such as the Kleene star. The treatment of the non-linear
types is pretty standard, so below we focus on the linear syntax.

@@

\paragraph{Linear Function Types}
The monoidal structure provided by $\otimes$ is both left and right closed and
this is denoted by the left and right linear function types. Further,
for linear types $A$ and $B$ the left linear function type $A \lto B$
enjoys an elimination principle similar to function application or modus ponens.
%
\[
\inferrule{\Delta \vdash A \otimes A \lto B}{\Delta \vdash B}
\]
%
The interpretation of linear function types at a string $w$ is a linear function
that takes in parses of $A$ on a string $w_{a}$ and outputs parses of $B$ on the
string $w_{a} w$.
%
\[
  \sem{A \lto B} \gamma w = \Pi(w_a:\String) \left( A \gamma w_a \Rightarrow B\gamma (w_a w) \right)
\]
%
That is, strings match $A \lto B$ if when prepended with a parse of $A$ they
complete to parses of $B$. In this manner, the linear function types generalize
Brzozowksi's notion of derivative
\cite{brzozowskiDerivativesRegularExpressions1964}.
Brzozowski initially only gave an accounting of this operation for
generalized regular expressions, but later work from Might et al.\ demonstrates that the same
construction can be generalized to context free grammars
\cite{mightParsingDerivativesFunctional2011}. Here, via the linear function
types, the same notion of derivative extends to the grammars of \theoryabbv.
Note, to ensure that the linear function types do indeed generalize Brzozowski
derivative, we must include the equalities in \cref{fig:brzozowskideriv} as axioms.
\pedro{We should further explain why these equations are required, or at the very
least give some intuition}

>>>>

\steven{The above paragraph maybe isn't the right thing to say, but I want
  something that ties the first paragraph to the contributions that follow}
\pedro{As a rule of thumb, I think that it's good to avoid subjective and
  pointed adjectives such as ``intuitive'', espcially if we don't have evidence
  that corroborates our claim. This is the perfect excuse for
an annoying reviewer to complain about our work.}
\max{I agree about ``intuitive'', try to stick to objective claims}

@@

for every string $w \in \String$. Intuitively, we read a morphism of grammars as
a \emph{parse transformer} --- i.e. $f$ translates a $G$-parse of $w$ into an $H$-parse
of $w$. The induced notion of isomorphism in $\Grammar$ encodes that two grammars are equivalent if there is
a bijective translation of parses. In other words, two grammars are isomorphic
if they capture the same structural descriptions of strings, which is
precisely Chomsky's notion of \emph{strong equivalence} of
grammars \cite{chom1963}. In works like
\cite{yoshinaga2002formal}, this notion of strong equivalence is often treated
like the appropriate choice of ``sameness'' for two grammars. When abstracting over the latent
structure of formal grammars and describing it in the the language of
categories, it is encouraging to see the same definition of equivalence arise
naturally.

Upon inspection of the above definition, $\Grammar$ may equivalently
be described as the functor category $\Set^{\String}$ --- where the
set $\String$ is viewed as a discrete\footnote{Objects in $\String$
comprise the set of all strings, and the only morphisms are the
identity morphisms.}  category. In fact, in this paper we will take
$\Grammar$ to be defined as the functor category
$\Set^{\String}$. Such functor categories into $\Set$ --- often called
presheaf categories --- carry a remarkable amount of structure, and we
will demonstrate that these well-known constructions on presheaf
categories provide us with familiar compositional constructions on
grammars\max{where do we talk about the properties of this presheaf cat?}.

@@

\steven{I initially wanted to say ``universal semantic domain to reason about
  parsing'', but these words have specific domain theory usage that aren't
  appropriate here, right?}
\pedro{If you want to keep ``semantic domain'', we could go for ``all-encompassing semantic domain''}
\steven{Check details for 2-cat of grammar formalisms}
\max{this section kind of trails off at the end with a very vague aside about a $2$-category of grammar formalisms. Maybe this should go in future work/discussion? What is the message of this section? How does it tie in with the message of the paper? What even is the message of this paper?}

@@

Omission of the structural rules of a deductive system, such as in
linear logic \cite{GIRARD19871}, offers precise control over how a value is used
in a derivation. Namely\max{namely?}, linear logic omits the weakening and contraction rules
to ensure that every value in context is used exactly once. This control enables
\emph{resource-sensitive} reasoning, where we may treat a resource as
\emph{consumed} after usage. This viewpoint is amenable to parsing applications,
as we may treat characters of a string as finite resources that are consumed at
parse-time. That is, once a particular substring has been parsed, a parser need not go
over it again\max{what about a backtracking parser?}. One may then envision a linear type system where the types comprise
formal grammars generated over some alphabet $\Sigma$, and the type constructors
correspond precisely to inductive constructions on grammars --- such as
conjunction, disjunction, concatenation, etc.

Unfortunately, programming in a linear term language is often
cumbersome and unintuitive for users.\max{You are arguing against our own type theory here...} Code in such a language can
become unnecessarily verbose when using Girard encodings to translate
high-level reasoning into low-level linear terms.\pedro{What encodings are these?
We should add a citation here}
To alleviate this
concern, in 1995 Benton et al.\ proposed an alternative categorical
semantics of linear logic based on adjoint interactions between linear
and non-linear logics \cite{bentonMixedLinearNonlinear1995} ---
appropriately referred to as a \emph{linear-non-linear} (LNL)
system. This work is simply typed, so the boundary between linear and
non-linear subtheories is entirely characterized via a monoidal
adjunction $F \dashv G$ between linear terms and non-linear terms.

@@

The structural judgments are shown in \cref{fig:structjdg}, the typing
well-formedness in \cref{fig:typewf}, and the intuitionistic typing
rules in \cref{fig:inttyping}. These are mostly just as they appear in
\cite{krishnaswami_integrating_2015}. It has two main differences. The
first one it the presence of two-distinct arrow types, one for adding
variables to the beginning of contexts and another one for adding
variables to the end of contexts. This is an adequate change in the
context of grammars, which are inherently non-symmetric. The second
change is the introduction of inductive types, which also have a
natural grammar counterpart\max{what is it then?}, which usually allows kinds of recursive
behavior such as the Kleene star. The treatment of the non-linear
types is pretty\max{pretty?} standard, so below we focus on the linear syntax.

@@

\paragraph{Linear Function Types}
The monoidal structure provided by $\otimes$ is both left and right closed and
this is denoted by the left and right linear function types. Further,
for linear types $A$ and $B$ the left linear function type $A \lto B$
enjoys an elimination principle similar to function application or modus ponens.
%
\[
\inferrule{\Delta \vdash A \otimes A \lto B}{\Delta \vdash B}
\]
%
The interpretation of linear function types at a string $w$ is a linear function
that takes in parses of $A$ on a string $w_{a}$ and outputs parses of $B$ on the
string $w_{a} w$.
%
\[
  \sem{A \lto B} \gamma w = \Pi(w_a:\String) \left( A \gamma w_a \Rightarrow B\gamma (w_a w) \right)
\]
%
That is, strings match $A \lto B$ if when prepended with a parse of $A$ they
complete to parses of $B$. In this manner, the linear function types generalize
Brzozowksi's notion of derivative
\cite{brzozowskiDerivativesRegularExpressions1964}.
Brzozowski initially only gave an accounting of this operation for
generalized regular expressions, but later work from Might et al.\ demonstrates that the same
construction can be generalized to context free grammars
\cite{mightParsingDerivativesFunctional2011}. Here, via the linear function
types, the same notion of derivative extends to the grammars of \theoryabbv.
Note, to ensure that the linear function types do indeed generalize Brzozowski
derivative, we must include the equalities in \cref{fig:brzozowskideriv} as axioms\max{we must? why must we}.
\pedro{We should further explain why these equations are required, or at the very
least give some intuition}
